---
title: "The mechanistic simulation benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The mechanistic simulation benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What the benchmark is for

The composite score is a ranking rule, so the only way to characterise it before
applying it to real data is to score datasets whose reservoir is known by
construction. `simulate_spillover()` generates such datasets: a stochastic SIRS
dynamic for one maintenance host, spillover-driven dynamics for the other
candidates, an optional bridge host, and an individual-level human population
with declared exposures, antibody waning and observation error.

`scenario_params()` returns the parameter list for each of the eleven scenarios
used in the paper: ten positive scenarios spanning the failure modes anticipated
from the conceptual literature, and one no-spillover negative control.

```{r scenarios, eval = FALSE}
library(spilloverID)
names(formals(scenario_params))
p <- scenario_params("S2_bridge")
str(p[c("n_sites", "n_hosts", "bridge_host", "alpha_reservoir", "alpha_bridge")])
```

## One replicate

```{r one, eval = FALSE}
sim <- simulate_spillover(scenario_params("S1_canonical"), seed = 20260524 + 17)
attr(sim, "truth")$k_star      # the host the simulator designated as reservoir
res <- compute_pi(sim)
res$pi
res$rank
round(res$criteria, 3)         # per-host, per-criterion scores
res$weights_used               # after redistribution of any NA criterion
```

The truth attribute is what makes the dataset a benchmark: `k_star` is the index
of the maintenance host, and `NA` in the negative control, where no host
maintains the pathogen.

## Scoring a scenario

Identification is scored two ways. Top-1 is a strict rank-1 call; the
discrimination margin is the score of the true reservoir minus the highest score
among the remaining candidates, and is negative whenever the call is wrong.

```{r loop, eval = FALSE}
scen <- "S1_canonical"
out <- do.call(rbind, lapply(1:20, function(rep) {
  d <- simulate_spillover(scenario_params(scen),
                          seed = 20260524 + 17 * rep + 1)
  k <- attr(d, "truth")$k_star
  r <- compute_pi(d)
  data.frame(rep = rep,
             top1   = as.integer(r$rank[k] == 1),
             margin = r$pi[k] - max(r$pi[-k]))
}))
mean(out$top1); mean(out$margin)
```

Two properties of the generative model should be kept in mind when reading such
numbers. First, only the true reservoir carries the transmission loop, so its
pooled PCR prevalence is the highest in essentially every replicate: the
benchmark contains no configuration in which an amplifier out-prevalences the
reservoir, which is precisely the inferential problem the composite exists to
address. Second, the inter-epidemic contrast that C12 reads saturates at its
ceiling for every host in most replicates, so the criterion carrying the largest
weight is unidentifiable there. A high simulated identification rate is
therefore a robustness check, not the principal validation.

## Reproducing the published benchmark

`run_benchmark.R` in the deposited `code/` directory runs all eleven scenarios
with the seeds fixed in the script and writes `benchmark_results.json` together
with the per-replicate criterion matrix and the reference rankings. It is
checkpointed per scenario, so it can be interrupted and restarted.
