---
title: "The three outbreak reconstructions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The three outbreak reconstructions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What the reconstructions are, and are not

Three outbreaks whose wildlife reservoir was later established by molecular work
are reconstructed from the summary values reported in the primary literature:
Sin Nombre hantavirus (southwestern United States, 1993), Nipah virus
(peninsular Malaysia, 1998–99) and Rift Valley fever virus (Lake Victoria basin,
2006–07).

These are simulated datasets anchored to published summary values, not primary
field data. Site-level seroprevalences, host-specific exposure prevalences,
exposure relative risks and inter-epidemic to epidemic ratios are set to the
published figures; individual records are then drawn under the catalytic model
so as to reproduce them. The intervals the framework returns therefore propagate
the sampling error of the reconstruction only, not the uncertainty in the
literature values used to parameterise it, which is likely to dominate. Every
reported interval should be read as a lower bound on the true uncertainty.

## Building a case dataset

The constructors live in `run_cases.R` in the deposited `code/` directory rather
than in the package, because they encode the published parameters of three
specific outbreaks rather than a general method. The data object they return is
an ordinary `spillover_data`.

```{r build, eval = FALSE}
library(spilloverID)
source("run_cases.R")          # defines case_sin_nombre(), case_nipah(), case_rvf()
set.seed(20260524)
d_sn <- case_sin_nombre()
d_sn
```

## Scoring, with the bootstrap

```{r score, eval = FALSE}
res <- compute_pi(d_sn, bootstrap = 100)
res$pi
res$rank
round(res$criteria, 3)
apply(res$bootstrap, 2, quantile, c(0.025, 0.975))
```

## The discrimination margin and its interval

The margin is the score of the host under consideration minus the highest score
among the remaining candidates. Its interval is the percentile interval of that
difference across the same bootstrap replicates, computed replicate by
replicate. Differencing the two per-host intervals instead would ignore the
correlation between hosts within a replicate and would overstate the width.

```{r margin, eval = FALSE}
k  <- which(names(res$pi) == "P. maniculatus")
bm <- res$bootstrap
d  <- bm[, k] - apply(bm[, -k, drop = FALSE], 1, max)
c(point = res$pi[k] - max(res$pi[-k]), quantile(d, c(0.025, 0.975)))
```

## Reading the result

Three things are worth checking before a ranking is acted on. Whether the
highest score reaches the interpretation threshold recalibrated for the set of
criteria actually available, since the scale of the score depends on how many
criteria discriminate. Whether the margin interval excludes zero. And whether
the C12 ranking agrees with the overall ranking: a divergence between the two is
the signature of a bridge-host or vector-borne configuration, in which the host
producing the proximate epidemiological signal is not the host that maintains
the pathogen.

```{r weights, eval = FALSE}
res$weights_used[res$weights_used > 0]   # criteria that were actually evaluable
res$criteria[, "C12"]
```
