---
title: "Getting started with spilloverID"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with spilloverID}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Purpose

`spilloverID` quantifies which candidate wildlife host is the most plausible
source of zoonotic spillover when only **serological and exposure data** are
available (no genetic data). It implements a composite epidemiological
plausibility score Π_k for each candidate host *k*, based on 14 independent
signatures, three of which (C12–C14) target the distinction between a
maintenance reservoir and an amplifying host.

## Required data structure

Three data frames at minimum:

### Human survey (`human`)

| site | ind | age | sex | prof | residence | sero | exp_h1 | exp_h2 | ... |
|------|-----|-----|-----|------|-----------|------|--------|--------|-----|
| 1    | 1   | 32  | 0   | 1    | 22        | 1    | 1      | 0      | ... |
| 1    | 2   | 47  | 1   | 0    | 47        | 0    | 0      | 0      | ... |

- `sero`: binary, result of serological assay.
- `exp_hk`: binary, declared exposure to host *k* (one column per host).
- Optional but recommended: `sex`, `prof` (occupational risk), `residence`
  (years at site), `household`, `env_declared`.

### Animal aggregations (`animal`)

| site | host | n_sampled | n_pcr | prev_pcr | n_sero | prev_sero |
|------|------|-----------|-------|----------|--------|-----------|
| 1    | 1    | 40        | 24    | 0.60     | 32     | 0.80      |
| 1    | 2    | 40        | 2     | 0.05     | 8      | 0.20      |

### Site covariates (`site`, optional)

| site | anthrop | [other env. variables] |
|------|---------|-------------------------|
| 1    | 0.42    | ...                     |

## Minimal usage

```{r minimal, eval = FALSE}
library(spilloverID)
data <- spillover_data(human, animal, site = site,
                       n_hosts = 4,
                       host_names = c("Oryzomys", "Peromyscus", "Heteromys", "Mus"))
result <- compute_pi(data)
print(result)
```

## With bootstrap confidence intervals

```{r boot, eval = FALSE}
result <- compute_pi(data, bootstrap = 500)
```

## Adapting weights

```{r weights, eval = FALSE}
# Mechanistic emphasis
w <- default_weights()
w["C6"] <- 0.20; w["C10"] <- 0.20
w <- w / sum(w)
result_mech <- compute_pi(data, weights = w)
```

## Interpreting Π_k

Indicative thresholds:

- **Π_k > 0.70** and ≥ 0.15 above second highest → highly plausible reservoir.
- **Π_k > 0.50** without clear dominance → multi-host system plausible.
- **Π_k < 0.30** → unlikely.

Always inspect the per-criterion matrix (`result$criteria`) to understand
which signatures are driving the score and whether convergence is achieved
across at least four of the seven catégories (spatial, individual,
multi-level, demographic, temporal, mechanistic, specificity).
